---
title: "data sampling"
output: html_document
---

In today's notes, we gave an example in which we discuss sampling data from the pdf:
\begin{align*}
f_X(x) = \frac{3}{8} x^2 ~~~ x \in [0,2] \,.
\end{align*}
Plotted, the pdf looks like this:

```{r fig.align='center',fig.width=3,fig.height=3}
x   <- seq(0,2,by=0.01)
f.X <- 3*x^2/8
library(ggplot2)
ggplot(data=data.frame(x,f.X),mapping=aes(x=x,y=f.X)) +
  geom_line(col="red") +
  ylab(expression(f[X]*"(x)"))
```

## Inverse-Transform Sampling

In the notes, we determined that the inverse cdf is
\begin{align*}
x = 2q^{1/3}
\end{align*}
given input $q \in [0,1]$. Let's use this to sample 1000 data from $f_X(x)$:

```{r fig.align='center',fig.width=3,fig.height=3}
set.seed(36235)
n        <- 1000
q        <- runif(n)  # by default, this samples between 0 and 1
x.sample <- 2*q^(1/3)

ggplot(data=data.frame(x.sample),mapping=aes(x=x.sample)) +
  geom_histogram(mapping=aes(y=after_stat(density)),bins=20,fill="blue") +
  geom_line(data=data.frame(x,f.X),mapping=aes(x=x,y=f.X),col="red") +
  ylab(expression(f[X]*"(x)"))
```

We can see that the histogram of sampled data values follows $f_X(x)$ well, while noting that the defined bin boundaries are such that the last bin only contains have the expected density of counts. (That can be fixed!)

## Rejection Sampling

When we examine the plot above, we see that the domain of $f_X(x)$ is $[0,2]$, while the maximum value of $f_X(x)$ is 1.5 (at $x=2$). We will use this information to code our rejection sampler, as per the algorithm given in the notes.

```{r fig.align='center',fig.width=3,fig.height=3}
set.seed(36235)
n        <- 1000
x.sample <- rep(NA,n)  # set aside storage for x
x.fail   <- c()   # we only include this here for plotting, do not include this normally
y.sample <- rep(NA,n)
y.fail   <- c()

ii <- 1
while ( is.na(x.sample[n]) == TRUE ) {  # basically, while we haven't filled in the last element of x
  u <- runif(1,min=0,max=2)   # sample uniformly from the domain
  v <- runif(1,min=0,max=1.5) # sample uniformly from the range
  if ( v <= 3*u^2/8 ) {
    x.sample[ii] <- u
    y.sample[ii] <- v
    ii <- ii+1
  } else {
    x.fail <- append(x.fail,u) # again, we would not have this else statement normally
    y.fail <- append(y.fail,v)
  }
}

ggplot(data=data.frame(x.sample),mapping=aes(x=x.sample)) +
  geom_histogram(mapping=aes(y=after_stat(density)),bins=20,fill="blue") +
  geom_line(data=data.frame(x,f.X),mapping=aes(x=x,y=f.X),col="red") +
  ylab(expression(f[X]*"(x)"))
```

The plot below illustrates the rejection algorithm well: we are simply sampling points uniformly within a rectangle, and keeping those data that "lie below" $f_X(x)$. (Our data sample is the $x$-coordinates of all the green points.) The algorithm is inefficient, however: just over 2/3rds of all candidate data get thrown out!

```{r fig.align='center',fig.width=3,fig.height=3}
ggplot(data=data.frame(x,f.X),mapping=aes(x=x,y=f.X)) +
  geom_line(col="red") +
  geom_point(data=data.frame(x.sample,y.sample),mapping=aes(x=x.sample,y=y.sample),col="green",cex=0.5) +
  geom_point(data=data.frame(x.fail,y.fail),mapping=aes(x=x.fail,y=y.fail),col="red",cex=0.5) +
  ylab(expression(f[X]*"(x)"))

cat("The number of candidate points is ",n+length(x.fail),".\n")
```
